---
title: "Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis"
author:
  - Ying Jin
  - Andrew Leroux
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
fontfamily: mathpazo
fontsize: 11pt
geometry: margin = 1in
bibliography: refs.bib
nocite: '@*'
link-citations: yes
linkcolor: blue
csl: ASA.cls
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage[utf8]{inputenc}
  - \usepackage[T1]{fontenc}
---
<!-- Abstract:  -->
<!-- ================================================================================================================================= -->
<!--   Biomedical investigators are often interested in predicting future observations of subjects based on their historical data, referred to as dynamic prediction. Traditional methods are often limited in flexibility and computationally intensive, especially with non-Gaussian data. To address these issues, we propose a novel method for dynamic prediction based on Generalized Functional Principal Component Analysis (FPCA). Assume the observed outcome follows an exponential family distribution parameterized by a latent Gaussian function, the proposed method consists of the following steps: 1) Bin the data across functional domain into small, equal-length intervals; 2) Fit local generalized mixed models at every bin to estimate individual latent functions; 3) Fit FPCA model to smooth latent functions and 4) Obtain estimates of subject-specific PC scores using partial observations and recover the unobserved part on the binned grid. Our simulation study showed the proposed method achieved significantly better out-of-sample predictive performance compared to existing methods with much shorter computation time, thus has the potential to be widely applicable to large datasets. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
theme_set(theme_minimal())
library(here)
library(ROCR)
library(kableExtra)
library(knitr)
library(ggpubr)
options(knitr.kable.NA = "")

set.seed(314)

# simulation
load(here("Data/sim_data.RData"))
load(here("Data/SimOutput_fGFPCA.RData"))
load(here("Data/SimOutput_GLMMadaptive.RData"))

# data application
df_appl <- read.csv(here("Data/Daily_Weigh_ins.csv"))
load(here("Data/ApplOutput_fGFPCA.RData"))
load(here("Data/ApplOutput_GLMMadaptive.RData"))

```


Introduction
=================================================================================================================================

Biomedical investigators are often interested in predicting future observations of subjects based on their historical data, typically referred to as dynamic prediction. Traditionally, this type of data has been modeled using marginal models (generalized estimating equations) or conditional models (mixed effect models) [@liang1986; @Laird1982; @lindstrom1990; @davidian2003], and predictions are made based on the correlation between repeated measures from the same subject, and/or covariates that can be either fixed or time-varying. However, these methods are limited in terms of flexibility of correlation structure and the ability to handle out-of-sample prediction. When sample size is large or the density of repeated measures is high, they also tend to cause severe computational burden [@GLMMadaptive]. 

To address these problems, one may turn to functional mixed effect models instead when measures are dense across the domain of measures. Such methods accommodate more flexible correlation structure by modeling subject-specific random effects as a function, and non-parametric smoothing [@Scheipl2014] can be incorporated to speed up the computation, such as spline basis functions or eigenfunctions from functional principal component analysis (FPCA). The introduction of basis functions also makes out-of-sample prediction more straightforward. Instead of estimating subject-specific random effects of new observations, we can simply estimate coefficients/loadings on the basis function used for smoothing. Existing research on dynamic prediction with functional data analysis methods has been focusing on continuous/Gaussian outcomes, modelling subject-specific random effects with FPCA [@chiou2012; @goldberg2014; @shang2017]. @kraus2015 has used this approach to predict missing observations in partially observed function tracks, and @delaigo2016 achieved similar goals using Markov Chains. While methods mentioned above used only partial observations for prediction with an intercept-only model, @leroux2016 proposed Functional Concurrent Regression (FCR) framework which can incorporate the effect of subject-specific predictors. 

Unfortunately, fewer papers have focused on its extension to non-Gaussian data, such as series of binary or count outcomes. Existing methods also tend to be very computationally intensive. For example, @chen2013 proposed approaches to fit marginal functional models that is compatible to multi-level, generalized outcomes. @goldsmith2015 established a model framework that takes into account the fixed effect of time-invariant covariates, with parameters estimated with Bayesian method in *Stan*. @gertheiss2016 identified bias introduced by directly applying FPCA methods to generalized functions, and proposed to address this problem using a two-stage, joint estimation strategy. @linde2019 used an adapted Bayesian variational algorithm for FPCA of binary and count data. In terms of implementation, @wrobel2019 proposed a fast, efficient way to fit GFPCA on binary data using EM algorithm, accompanied by the an open source R package *registr*. 

In this paper, we aim to develop a fast, scalable method for dynamic prediction of discrete function tracks based on functional mixed effect model with FPCA smoothing. Section 2 presents the procedure of the proposed method. In Section 3, we illustrate the performance and efficiency of our proposed method in a simulation study. In Section 4, we apply this method to a real-world dataset. Section 5 presents a discussion of advantages and limitation of the proposed method. 



Method
=================================================================================================================================

The observed data for a single subject $i$ is ($t$, $Y_i(t)$), where $t$ consists of dense, discrete points along the functional domain, and $Y_i(t)$ is the non-Gaussian outcome observed at $t$. We assume that the outcome $Y_i(t)$ can be characterized by a latent continuous function $\eta_i(t)$. That is, at a specific t, $Y_i(t)$ follows a exponential family distribution such that:

$$
g[E(Y_i(t))] = \eta_i(t) = \beta_0(t)+b_i(t)
$$


where g is an appropriate link function, $\beta_0(t)$ is the population mean of latent function, and $b_i(t)$ is a subjects-level random effect function following a zero-mean Gaussian process. 

According to the Karhunen-Loève theorem, the unobserved random effect function $b_i(t)$ can be represented as the linear combination of a infinite set of orthogonal functions: $b_i(t)=\sum_{k=1}^{\infty}\xi_{ik}\phi_{k}(t)$. We approximate this function by a finite set of $\phi_{k}(t)$, such that $b_i(t)=\sum_{k=1}^{K}\xi_{ik}\phi_{k}(t)+\epsilon_i(t)$, where $\epsilon_i(t)$ indicates random noise introduced by this approximation, following a zero-mean Gaussian process. At each specific point t, $\epsilon_i(t) \sim N(0, \sigma^2)$.

In practice, this approximation could be implemented through an FPCA procedure on $\eta_i(t)$. In this case $\phi_{k}(t),\ k \in \{1,...,K\}$ are the first K eigenfunctions that explain the most variation in $\eta_i(t)$, and $\xi_{ik}$ are subject-specific PC scores (or loadings) on each eigenfunction. Additionally, $\xi_{ik}$ are mutually independent over both subject ($i$) and eigenfunctions ($k$). That is, each $\xi_{ik}$ follows normal distribution $N(0, \lambda_k)$ where $\lambda_k$ is the kth eigenvalues.


**Fast Generalized FPCA** Based on the problem set up above, we propose the following algorithm for fast, generalizable implementation of FPCA (fGFPCA) on the unobserved latent process $\eta_i(t)$:

1. Bin the observed outcomes in to small, non-overlapping, equal length intervals. We hereafter index the bins by their midpoints $s$. 
2. Fit a local Generalized Mixed Model at every bin. Specifically, at bin $s$, we fit $g[E(Y_i(t))] = \beta_0(s)+b_i(s)$ for all $t$ in bin $s$. From this series of models we can estimate the individual latent functions at every bin: $\hat{\eta}_i(s) = \hat{\beta}_0(s)+\hat{b}_i(s)$. 
3. Fit FPCA on the estimated latent functions: $\hat{\eta}_i(s) = f_0(s)+\sum_{k=1}^K\xi_{ik}\phi_{k}(s)+\epsilon_i(s)$, and obtain estimates of basis functions $\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}$, eigenvalues $\hat{\lambda}_1...\hat{\lambda}_k$, mean function $\hat{f}_0(s)$ and residual variance $\hat{\sigma}^2$. 


**Dynamic prediction** Now assume we have a new observations with partially observed outcome $Y_{new}(t)$, $t\leq t_m$. The maximum observed point is $t_m$, which is a point along the functional domain and belongs to bin $s_m$ on the binned grid from step 1 above. With components extracted from the FPCA model in step 3, we will be able to make predictions on the future, unobserved bins. Specifically, with the partially observed binary function $Y_{new}(t)$, we will be able to calculate the maximum likelihood estimate (MLE) of the subject-specific PC scores $\hat{\xi}_{ik}$. Then the value of latent functions at unobserved points can be estimated as $\hat{\eta_i}(s)=\hat{f}_0(s)+\sum_{k=1}^K\hat{\xi}_{ik}{\phi}_k(s)$

Following the algorithm above, predictions of individual latent functions are made on the binned grid based on partially observed non-Gaussian functions tracks of new subjects. Since the bins are set up to be small in length, the binned grid would still be dense. However, in situations where predictions on the original, un-binned grid is needed, linear interpolation turns out to be a fast, convenient way with good performance for prediction at points between bins. 

<!-- Precision of prediction, usually quantified by the variance of prediction error $Var(\hat{\eta_i}(s)-\eta_i(s))$, is also straightforward under this framework. In step 4 we calculated the MLE of $\hat{\xi}_{ik}$. Based on likelihood theory, its variance can be estimated with observed information ${I}(\hat{\xi}_{ik})$, which is essentially the second derivative of likelihood at  $\hat{\xi}_{ik}$. Therefore, the variation of prediction interval is: -->

<!-- $$Var(\hat{\eta_i}(s)-\eta_i(s))=\boldsymbol{\Phi}(s)I(\hat{\boldsymbol{\xi}}_i)\boldsymbol{\Phi}^T(s)+\hat{\sigma_{\epsilon}}^2$$ -->
<!-- Where $\boldsymbol{\Phi}(s)=(\phi_1(s)...\phi_K(s))$ and $\hat{\boldsymbol{\xi}}_i=(\hat{\xi}_{i1}, ...,\hat{\xi}_{iK})$. -->


<!-- - Add section of de-bias score: conditional on subject. Need repeat simulation to demonstrate -->

Simulation
=================================================================================================================================

In this section, we illustrate the predictive performance and computational efficiency of the proposed method through a simulation study. We simulated 50 datasets, each with 500 subjects. For every subject, we generate 1000 binary outcomes $Y_i(t) \in \{0, 1\}$ across functional domain $t \in [0, 1000]$, where the distribution of outcome is characterized by a continuous latent function. The data generation mechanism can be expressed as follows: 

$$\begin{aligned}
Y_i(t) & \sim Binomial(\frac{exp(\eta_i(t))}{1+exp(\eta_i(t))}) \\
\eta_i(t) &= f_0(t)+ \xi_{i1}\sqrt{2}sin(2\pi t)+\xi_{i2}\sqrt{2}cos(2\pi t)+\xi_{i3}\sqrt{2}sin(4\pi t)+\xi_{i4}\sqrt{2}cos(4\pi t)
\end{aligned}$$

In this simulation, we set $f_0(t) = 0$. $\xi_{ik}$ are mutually independent normal random variables $\xi_{ik}\sim N(0, \gamma_k)$. Here we set the values of $\gamma_k$ to be $0.5^{k-1}$, $k \in (1,..., 4)$. In addition, for simplicity of presentation, we generate data on a regular grid, which means observations points are equally distributed across $[0, 1000]$ and are the same for all subjects. 

We use two metrics to evaluate the out-of-sample predictive performance: integrated squared error (ISE) and Area-Under-the-Receiver-Operator-Curve (AUC). ISE assess the prediction accuracy of latent continuous function. It is evaluated on the binned grid at midpoints of each unobserved bin. If the entire functional domain has S bins, but we have observations up to the mth bin, then ISE is defined as $\frac{1}{N}\sum_{i=1}^N\sum_{s=s_m+1}^S(\hat{\eta}_i(s)-\eta_i(s))^2$. The second metric, AUC, focuses on evaluation of prediction of the binary outcome. Since the binary outcomes are generated on the original, un-binned grid, we evaluated AUC on this grid and estimated values of latent functions between bins with linear interpolation. 

As a reference method, we compare our method to Generalized Linear Mixed Models using Adaptive Gaussian Quadrature (GLMMadaptive). This is one of the fastest existing method developed for dynamic prediction of repeated generalized outcomes. Just like many mixed models, this method is very limited in terms of flexibility. For example, the model used for prediction of our simulated datset would simply be an linear model with one covariate indicating observation time: $g(E(Y_i)) = \beta_0+\beta_1t+b_{i0}+b_{i1}t$. While the flexibility of this mixed model can be increased using spline functions, the dimension of spline basis is also restricted by computational ability, and is unfeasible to implement under the scale of our simulated data or the complexity of our proposed method.

The average ISE and AUC across all simulation is presented in Table 1, and Figure 1 presents the predicted function curves of four randomly-drawn subjects from the first simulation. The prediction is conditioning on different length of observed track (specifically, with observations up to t = 200, 400, 600, 800 respectively), and evaluation is made on equal-length time windows on the unobserved tracks following the maximum observation time. As Table 1 reveals, our proposed method (fGFPCA) outperformed GLMMadaptive under every scenario, which was expected since GLMMadaptive can only accommodate simple model structures. In addition, fGFPCA also took less computation time. For one simulated dataset, fGFPCA spent `r round(mean(runtime),2)` minutes on model fitting and out-of-sample prediction, while GLMMadaptive took `r round(mean(runtime_ref),2)` minutes. 

Because of the flexibility of our proposed model framework, the accuracy of prediction at specific time points would improve with more observed data. This is revealed in Table 1 as ISE decreases and AUC increases with maximum observed time (left-to-right), also in Figure 1 as predicted curves get closer with longer observed track. However the same tendency is not observed with GLMMadaptive models, as a result of restricted model flexibility. A linear model that fits well to a specific part of latent function can fit very badly to the following parts, especially when the underlying latent function has cyclic patterns. Therefore, more observations do not necessarily make the model more predictive. 


```{r}
# other quantities
mid <- unique(pred_lst[[1]]$bin)
N <- 500
M <- length(pred_lst)
J <- 1000

# for fGFPCA
# code prediction value at observed points as NA
for(m in 1:length(pred_lst)){
  pred_lst[[m]]$pred_t200[pred_lst[[m]]$bin <= 200] <- NA
  pred_lst[[m]]$pred_t400[pred_lst[[m]]$bin <= 400] <- NA
  pred_lst[[m]]$pred_t600[pred_lst[[m]]$bin <= 600] <- NA
  pred_lst[[m]]$pred_t800[pred_lst[[m]]$bin <= 800] <- NA
}

# for GLMM adaptive
# add in observation time index
for(m in 1:length(pred_lst_ref)){
  pred_lst_ref[[m]]$sind_inx <- rep(1:J, N)
}
```


```{r Figure, fig.cap="Predicted track of four randomly selected subjects from one simulated dataset. The grey solid line indicates the true latent continuous function. The dashed lines indicated predicted latent function tracks, and color indicates different observation time.", fig.height=6, fig.width=12}
# figure of fGFPCA prediction
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# true data
df_true_sim1 <- sim_data[[1]] %>% select(id, sind_inx, eta_i)

# prediction results used for visualization
exp_df_sim1 <- pred_lst[[1]]

exp_df_sim1_ref <- pred_lst_ref[[1]]

# randomly draw four samples for visualization
rand_id <- sample(exp_df_sim1$id, 4)

# fGFPCA
fig1a <- exp_df_sim1 %>%
  left_join(df_true_sim1 %>% filter(sind_inx %in% mid) %>% rename(bin=sind_inx), by = c("id", "bin")) %>%
  filter(id %in% rand_id) %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_i, col = "True"))+
  geom_line(aes(x=bin, y = pred_t200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t400, col = "Up to 400"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t600, col = "Up to 600"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=bin, y = pred_t800, col= "Up to 800"), linetype = "dashed", na.rm = T)+
  facet_wrap(~id, scales = "free_y")+
  labs(y = "Latent function", x = "Bin", colour= "Observed time", title = "fGFPCA")+
  scale_color_manual(values=cbPalette)+
  scale_x_continuous(breaks = c(0, 200, 400, 600, 800, 1000))

# GLMMadaptive
fig1b <- exp_df_sim1_ref %>%
  filter(sind_inx %in% mid) %>%
  filter(id %in% rand_id) %>%
  ggplot()+
  geom_line(aes(x=sind_inx, y=eta_i, col = "True"))+
  geom_line(aes(x=sind_inx, y = pred_t200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_t400, col = "Up to 400"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_t600, col = "Up to 600"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_t800, col= "Up to 800"), linetype = "dashed", na.rm = T)+
  facet_wrap(~id, scales = "free_y")+
  labs(y = "Latent function", x = "Bin", colour= "Observed time", title = "GLMMadaptive")+
  scale_color_manual(values=cbPalette)+
  scale_x_continuous(breaks = c(0, 200, 400, 600, 800, 1000))

ggarrange(fig1a, fig1b, nrow = 1, common.legend = T)


```


```{r ISE_fGFPCA}
# calculate ISE on binned grid
ise_mat <- array(NA, dim = c(5, 4, M)) # dimensions are: predicted time, max obs time, simulation

for(m in 1:M){
  
  # true value at midpoints
  true_eta <- sim_data[[m]] %>% filter(sind_inx %in% mid) %>% select(id, sind_inx, eta_i) %>% rename(bin=sind_inx)
  pred_eta <- pred_lst[[m]] %>% select(-eta_hat)
  pred_eta <- pred_eta %>% left_join(true_eta, by = c("id", "bin"))
  
  # split unobserved time into intervals of length 200
  # [0, 200], (200, 400], (400, 600], (600, 800], (800, 1000]
  pred_eta$bin_cat <- cut(pred_eta$bin, breaks=c(0, 200, 400, 600, 800, 1000), include.lowest = T)
  
  # SE at each observed time
  pred_eta <- pred_eta %>% mutate(ise_t200 = (pred_t200-eta_i)^2,
                                  ise_t400 = (pred_t400-eta_i)^2,
                                  ise_t600 = (pred_t600-eta_i)^2,
                                  ise_t800 = (pred_t800-eta_i)^2)
  
  # subject-average SE
  ise_mat[, , m] <- pred_eta %>% group_by(bin_cat) %>% 
    summarize_at(vars(starts_with("ise")), .fun=function(x)(sum(x)/N)) %>%select(-bin_cat) %>% as.matrix()
}

ave_ise_mat <- apply(ise_mat, MARGIN = c(1, 2), mean)[-1, ]
colnames(ave_ise_mat) <- c("200", "400", "600", "800")
rownames(ave_ise_mat) <- c("(200, 400]", "(400, 600]", "(600, 800]", "(800, 1000]")


```

```{r, ISE_GLMMadaptive}
# calculate ISE on binned grid
ise_mat_ref <- array(NA, dim = c(5, 4, M)) # dimensions are: predicted time, max obs time, simulation

for(m in 1:M){
  
  # true value at midpoints
  pred_eta <- pred_lst_ref[[m]] %>% filter(sind_inx %in% mid)
  
  # split unobserved time into intervals of length 200
  # [0, 200], (200, 400], (400, 600], (600, 800], (800, 1000]
  pred_eta$bin_cat <- cut(pred_eta$sind_inx, breaks=c(0, 200, 400, 600, 800, 1000), include.lowest = T)
  
  # SE at each observed time
  pred_eta <- pred_eta %>% mutate(ise_t200 = (pred_t200-eta_i)^2,
                                  ise_t400 = (pred_t400-eta_i)^2,
                                  ise_t600 = (pred_t600-eta_i)^2,
                                  ise_t800 = (pred_t800-eta_i)^2)
  
  # subject-average SE
  ise_mat_ref[, , m] <- pred_eta %>% group_by(bin_cat) %>% 
    summarize_at(vars(starts_with("ise")), .fun=function(x)(sum(x)/N)) %>%select(-bin_cat) %>% as.matrix()
  
}

ave_ise_mat_ref <- apply(ise_mat_ref, MARGIN = c(1, 2), mean)[-1, ]
colnames(ave_ise_mat_ref) <- c("200", "400", "600", "800")
rownames(ave_ise_mat_ref) <- c("(200, 400]", "(400, 600]", "(600, 800]", "(800, 1000]")

# cbind(ave_ise_mat, ave_ise_mat_ref)
```


```{r AUC_fGFPCA, cache=TRUE}
# maximum observations time
t_vec <- c(200, 400, 600, 800)

# calculate AUC on unbinned grid
auc_mat <- array(NA, dim = c(4, 4, M)) # row index iteration, col index max observation time

for(m in 1:M){
  
  Y_df <- sim_data[[m]]
  eta_pred <- pred_lst[[m]]
 
  # use interpolation to extend prediction to the unbinned grid
  for(i in 1:N){
   Y_df$pred_200[Y_df$id==i] <- approx(x=eta_pred$bin[eta_pred$id==i], y=eta_pred$pred_t200[eta_pred$id==i], xout = 1:J)$y
   Y_df$pred_400[Y_df$id==i] <- approx(x=eta_pred$bin[eta_pred$id==i], y=eta_pred$pred_t400[eta_pred$id==i], xout = 1:J)$y
   Y_df$pred_600[Y_df$id==i] <- approx(x=eta_pred$bin[eta_pred$id==i], y=eta_pred$pred_t600[eta_pred$id==i], xout = 1:J)$y
   Y_df$pred_800[Y_df$id==i] <- approx(x=eta_pred$bin[eta_pred$id==i], y=eta_pred$pred_t800[eta_pred$id==i], xout = 1:J)$y
  }
  
  # split unobserved time into intervals of length 200
  # [0, 200], (200, 400], (400, 600], (600, 800], (800, 1000]
  Y_df$bin_cat <- cut(Y_df$sind_inx, breaks=c(0, 200, 400, 600, 800, 1000), include.lowest = T)
  Y_df_lst <- split(Y_df, f = Y_df$bin_cat)
  Y_df_lst <- Y_df_lst[-1]
  
  for(j in seq_along(t_vec)){
    
    t <- t_vec[j]
    pred_name <- paste("pred_", t, sep = "")
    
    # calculate AUC
    auc_lst <- lapply(Y_df_lst[j:4], 
                       function(x){
                         row_id <- !is.na(x[, pred_name])
                         this_eta <- x[row_id, pred_name]
                         this_Y <- x[row_id, "Y"]
                         auc_pref <- performance(prediction(this_eta, this_Y), "auc")
                         return(auc_pref@y.values)
                })
      
    auc_vec <- c(rep(NA, 4-length(auc_lst)), unlist(auc_lst))
    auc_mat[ , j, m] <- auc_vec

}}


ave_auc_mat <- apply(auc_mat, MARGIN = c(1, 2), mean)
colnames(ave_auc_mat) <- c("200", "400", "600", "800")
rownames(ave_auc_mat) <- c("(200, 400]", "(400, 600]", "(600, 800]", "(800, 1000]")

```

```{r AUC_GLMMadaptive}
# calculate AUC on unbinned grid
auc_mat_ref <- array(NA, dim = c(4, 4, M)) # row index iteration, col index max observation time

for(m in 1:M){
  
  Y_df <- sim_data[[m]]
  eta_pred <- pred_lst_ref[[m]]
  
  Y_df <- Y_df %>% left_join(eta_pred %>% select(id, sind_inx, starts_with("pred_")), by = c("id", "sind_inx"))
 
  # split unobserved time into intervals of length 200
  # [0, 200], (200, 400], (400, 600], (600, 800], (800, 1000]
  Y_df$bin_cat <- cut(Y_df$sind_inx, breaks=c(0, 200, 400, 600, 800, 1000), include.lowest = T)
  Y_df_lst <- split(Y_df, f = Y_df$bin_cat)
  Y_df_lst <- Y_df_lst[-1]
  
  for(j in seq_along(t_vec)){
    
    t <- t_vec[j]
    pred_name <- paste("pred_t", t, sep = "")
    
    # calculate AUC
    auc_lst <- lapply(Y_df_lst[j:4], 
                       function(x){
                         row_id <- !is.na(x[, pred_name])
                         this_eta <- x[row_id, pred_name]
                         this_Y <- x[row_id, "Y"]
                         auc_pref <- performance(prediction(this_eta, this_Y), "auc")
                         return(auc_pref@y.values)
                })
      
    auc_vec <- c(rep(NA, 4-length(auc_lst)), unlist(auc_lst))
    auc_mat_ref[ , j, m] <- auc_vec

  }
}



ave_auc_mat_ref <- apply(auc_mat_ref, MARGIN = c(1, 2), mean)
colnames(ave_auc_mat_ref) <- c("200", "400", "600", "800")
rownames(ave_auc_mat_ref) <- c("(200, 400]", "(400, 600]", "(600, 800]", "(800, 1000]")


```

```{r format table}
ise_tb <- cbind(ave_ise_mat, ave_ise_mat_ref)
auc_tb <- cbind(ave_auc_mat, ave_auc_mat_ref)


rbind(ise_tb, auc_tb) %>%
  kable(booktabs = T, digit=2,
        caption = "Predictive performance of fGFPCA and GLMM adaptive on the simulated datasets. ISE and AUC are average values across all 50 simulations.",
        align = "lcccccccc") %>%
  kable_styling() %>%
  add_header_above(c(" " = 1, "fGFPCA" = 4, "GLMMadaptive" = 4)) %>%
  add_header_above(c(" " = 1, "Maximum observed time" = 8)) %>% 
  group_rows(index = c("Prediction time window" = 8)) %>%
  group_rows(index = c("ISE" = 4, "AUC" = 4)) 
```



<!-- -	Repeat simulation: repeat 10-100 times -->

<!-- -	Different set-up: -->
<!-- a.	Different eigenfunctions: with or without periodicity (start with the current one) -->
<!-- b.	Outcome: binary or count -->
<!-- c.	Sample size: start with N=500 -->
<!-- d.	Grid density: start with J=1000 -->
<!-- e.	Bin width, overlap or not: start with non-overlap, bin width = 10 (100 bins) -->

Data application
=================================================================================================================================


```{r}
# data clean
mid2 <- unique(df_est_latent$bin)
N2 <- max(df_est_latent$id)
J2 <- 400

# for fGFPCA
# code prediction value at observed points as NA
df_est_latent$pred_t100[df_est_latent$bin <= 100] <- NA
df_est_latent$pred_t200[df_est_latent$bin <= 200] <- NA
df_est_latent$pred_t300[df_est_latent$bin <= 300] <- NA


# for GLMM adaptive
# add in observation time index
for(m in 1:length(pred_lst_ref)){
  pred_lst_ref[[m]]$sind_inx <- rep(1:J, N)
}
```

```{r AUC_fGFPCA_Appl, cache=TRUE}
# maximum observations time
t_vec2 <- c(100, 200, 300)

# calculate AUC on unbinned grid
auc_mat2 <- matrix(NA, nrow = 3, ncol = 3) # row index iteration, col index max observation time

# result container for interpolation
Y_df2 <- df_appl %>% select(participant_id, starts_with("day")) %>% 
  pivot_longer(2:401, values_to = "Y", names_to = "sind_inx") %>%
  mutate(sind_inx = gsub("day_", "", sind_inx)) %>%
  mutate(sind_inx = as.numeric(sind_inx)) %>%
  rename("id" = "participant_id")
Y_df2$pred_100 <- Y_df2$pred_200 <- Y_df2$pred_300 <- NA

  # use interpolation to extend prediction to the unbinned grid
for(i in 1:N2){
   Y_df2$pred_100[Y_df2$id==i] <- approx(x=df_est_latent$bin[df_est_latent$id==i], 
                                         y=df_est_latent$pred_t100[df_est_latent$id==i], 
                                         xout = 1:J2)$y
   Y_df2$pred_200[Y_df2$id==i] <- approx(x=df_est_latent$bin[df_est_latent$id==i], 
                                         y=df_est_latent$pred_t200[df_est_latent$id==i], 
                                         xout = 1:J2)$y
   Y_df2$pred_300[Y_df2$id==i] <- approx(x=df_est_latent$bin[df_est_latent$id==i], 
                                         y=df_est_latent$pred_t300[df_est_latent$id==i], 
                                         xout = 1:J2)$y
  }
  
# split unobserved time into intervals of length 100
Y_df2$bin_cat <- cut(Y_df2$sind_inx, breaks=c(0, 100, 200, 300, 400), include.lowest = T)
Y_df_lst2 <- split(Y_df2, f = Y_df2$bin_cat)
Y_df_lst2 <- Y_df_lst2[-1]
  
for(j in seq_along(t_vec2)){
    
    t <- t_vec2[j]
    pred_name <- paste("pred_", t, sep = "")
    
    # calculate AUC
    auc_lst <- lapply(Y_df_lst2[j:3], 
                       function(x){
                         row_id <- !is.na(x[, pred_name])
                         this_eta <- x[row_id, pred_name]
                         this_Y <- x[row_id, "Y"]
                         auc_pref <- performance(prediction(this_eta, this_Y), "auc")
                         return(auc_pref@y.values)
                })
      
    auc_vec <- c(rep(NA, 3-length(auc_lst)), unlist(auc_lst))
    auc_mat2[ , j] <- auc_vec

}


colnames(auc_mat2) <- c("100", "200", "300")
rownames(auc_mat2) <- c("(100, 200]", "(200, 300]", "(300, 400]")

```

```{r AUC_GLMMadaptive_Appl}
# calculate AUC on unbinned grid
auc_mat_ref2 <- matrix(NA, nrow = 3, ncol = 3)  # row index iteration, col index max observation time

# split unobserved time into intervals of length 100
df_est_latent_ref$bin_cat <- cut(df_est_latent_ref$sind_inx, 
                                 breaks=c(0, 100, 200, 300, 400), include.lowest = T)
Y_df_lst2_ref <- split(df_est_latent_ref, f = df_est_latent_ref$bin_cat)
Y_df_lst2_ref <- Y_df_lst2_ref[-1]
  
for(j in seq_along(t_vec2)){
    
    t <- t_vec2[j]
    pred_name <- paste("pred_t", t, sep = "")
    
    # calculate AUC
    auc_lst <- lapply(Y_df_lst2_ref[j:3], 
                       function(x){
                         row_id <- !is.na(x[, pred_name])
                         this_eta <- x[row_id, pred_name]
                         this_Y <- x[row_id, "Y"]
                         auc_pref <- performance(prediction(this_eta, this_Y), "auc")
                         return(auc_pref@y.values)
                })
      
    auc_vec <- c(rep(NA, 3-length(auc_lst)), unlist(auc_lst))
    auc_mat_ref2[ , j] <- auc_vec

  }


colnames(auc_mat_ref2) <- c("100", "200", "300")
rownames(auc_mat_ref2) <- c("(100, 200]", "(200, 300]", "(300, 400]")


```


```{r, Figure_appl, fig.cap="Predicted track of four randomly selected subjects from the daily weigh-in dataset. The grey solid line indicates the true latent continuous function. The dashed lines indicated predicted latent function tracks, and color indicates different observation time.", fig.height=6, fig.width=12}
rand_id2 <- sample(N2, 4)
fig2a <- Y_df2 %>%
  filter(id %in% rand_id2) %>%
  mutate_at(vars(pred_100, pred_200, pred_300), function(x)exp(x)/(1+exp(x))) %>%
  ggplot()+
  geom_point(aes(x=sind_inx, y=Y, col = "True outcome"), size = 0.5)+
  geom_line(aes(x=sind_inx, y = pred_100, col = "Up to 100"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_300, col = "Up to 300"), linetype = "dashed", na.rm = T)+
  facet_wrap(~id, scales = "free_y")+
  labs(y = "Predicted latent function on probability scale", x = "Bin", 
       colour= "Observed time", title = "fGFPCA")+
  scale_color_manual(values=cbPalette)+
  scale_x_continuous(breaks = c(0, 100, 200, 300, 400))

fig2b <- df_est_latent_ref %>%
  filter(id %in% rand_id2) %>%
  mutate_at(vars(pred_t100, pred_t200, pred_t300), function(x)exp(x)/(1+exp(x))) %>%
  ggplot()+
  geom_point(aes(x=sind_inx, y=Y, col = "True outcome"), size = 0.5)+
  geom_line(aes(x=sind_inx, y = pred_t100, col = "Up to 100"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_t200, col = "Up to 200"), linetype = "dashed", na.rm = T)+
  geom_line(aes(x=sind_inx, y = pred_t300, col = "Up to 300"), linetype = "dashed", na.rm = T)+
  facet_wrap(~id, scales = "free_y")+
  labs(y = "Predicted latent function on probability scale", x = "Bin", 
       colour= "Observed time", title = "GLMMadaptive")+
  scale_color_manual(values=cbPalette)+
  scale_x_continuous(breaks = c(0, 100, 200, 300, 400))

ggarrange(fig2a, fig2b, nrow = 1, common.legend = T)
```

```{r}
cbind(auc_mat2, auc_mat_ref2) %>%
  kable(booktabs = T, digit=2,
        caption = "AUC of fGFPCA and GLMM adaptive on the daily weigh-in dataset",
        align = "lcccccccc") %>%
  kable_styling() %>%
  add_header_above(c(" " = 1, "fGFPCA" = 3, "GLMMadaptive" = 3)) %>%
  add_header_above(c(" " = 1, "Maximum observed time" = 6)) %>% 
  group_rows(index = c("Prediction time window" = 3))

```
Discussion
=================================================================================================================================
-	Grid 
-	Score bias: cannot demonstrate without repeat simulation

References
=================================================================================================================================
<div id="refs"></div>