---
title: "Problem of Step 4 (debias) of fGFPCA on NHANES data"
author: "Ying Jin"
output: 
  html_document:
    self_contained: yes
    number_sections: true
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: true
    font: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

set.seed(121)

library(here)
library(tidyverse)
library(ggpubr)
library(refund)
library(lme4)
library(ROCR)
library(GLMMadaptive)
library(kableExtra)
library(mvtnorm)
library(mgcv)
library(splines)
library(LaplacesDemon)
theme_set(theme_minimal())
```


```{r}
# load data
df <- read_rds(here("Data/nhanes_bi.rds"))

# code
source(here("Code/GLMM-FPCA.R")) 
# use pred_latent function to estimate latent function 
K <- 4
source(here("Code/OutsampBayes.R"))
```


Fit fGFPCA model

Because step 4 (debias) is very time-consuming, this example is fit on a subset of NHANES dataset, including 1000 participants as the training set

```{r data_subset}
# head(df)
df <- df %>% rename(id=SEQN, Y=Z)

Ntr <- 1000

train_id <- unique(df$id)[1:Ntr]
train_df <- df %>% filter(id %in% train_id)

# additional parameters
J <- length(unique(df$sind))
```

# Step 1: Bin data

- Bin every 10 observations

```{r, message=FALSE}
# bin data
bin_w <- 10 # bin width
n_bin <- J/bin_w # number of bins
brks <- seq(0, J, by = bin_w) # cutoff points
mid <- (brks+bin_w/2)[1:n_bin] # mid points

train_df$bin <- cut(train_df$sind, breaks = brks, include.lowest = T, labels = mid)
train_df$bin <- as.numeric(as.character(train_df$bin))
```


# Step 2:  Local GLMM

```{r}
# fit model on the training set
train_bin_lst <- split(train_df, f = train_df$bin)

t1=Sys.time()
df_est_latent <- lapply(train_bin_lst, function(x){pred_latent(x, n_node = 0)}) 
t2= Sys.time()
# t2-t1 
```


```{r}
df_est_latent <- bind_rows(df_est_latent) 
# head(df_est_latent)

# example estimated latent function
rand_id <- sample(train_id, 4)
df_est_latent %>% 
  filter(id %in% rand_id) %>%
  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  ggplot()+
  geom_line(aes(x=sind, y=eta_hat, group = id, col = "estimated"))+
  geom_point(aes(x=sind, y = Y, group = id, col = "outcome"), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  labs(x = "Time", y = "Estimated latent function (probablity scale)")
```

# Step 3: FPCA

```{r}
uni_eta_hat <- df_est_latent %>% filter(bin==sind)
mat_est_unique <- matrix(uni_eta_hat$eta_hat,
                         nrow=Ntr, 
                         ncol=n_bin, byrow = F) 
# row index subject, column binned time

t1 <- Sys.time()
fpca_mod <- fpca.face(mat_est_unique, argvals = mid, var=T, npc=K)
# keep 4 PCs
t2 <- Sys.time()
# t2-t1 
```


```{r PC}
data.frame(t = mid, fpca_mod$efunctions[, 1:4]) %>%
  rename(PC1 = X1, PC2=X2, PC3=X3,PX4=X4) %>%
  pivot_longer(2:5, names_to = "PC") %>%
  ggplot()+
  geom_line(aes(x=t, y=value, col = PC))+
  labs(x="Time", y="", title = "Eigenfunctions from FPCA on the binned grid")
```

- Eigenvalues: `r fpca_mod$evalues`
- Proportion of variance explained with four PCs: `r fpca_mod$pve`


# Step 4: Re-evaluation

## Extend the eigenfunctions back to the original grid with B-spline basis 

```{r interplate}
# Re-evaluation
## grid extension
p <- 3 # order of b splines 
knots <- 35 # number of knots (same from FPCA model)
knots_values <- seq(-p, knots + p, length = knots + 1 + 2 *p)/knots
knots_values <- knots_values * (max(mid) - min(mid)) + min(mid)

B <- spline.des(knots = knots_values, x = mid, ord = p + 1,
                outer.ok = TRUE)$design  # evaluate B-splines on binned grid
Bnew <- spline.des(knots = knots_values, x = 1:J, ord = p + 1,
                   outer.ok = TRUE)$design  # evaluate B-splines on original grid


df_phi <- matrix(NA, J, K) 
for(k in 1:K){
  lm_mod <- lm(fpca_mod$efunctions[,k] ~ B-1)
  df_phi[,k] <- Bnew %*% coef(lm_mod)
}# project binned eigenfunctions onto the original grid
```


Below presents the result of re-evaluation. Points are values of PC functions on the binned grid estimated from FPCA, and lines are values of re-evaluated PC functions on the original measurement grid. 

```{r}
df_pc1 <- data.frame(t=1:J,  df_phi) %>%
  rename(PC1=X1, PC2=X2, PC3=X3, PC4=X4) %>%
  pivot_longer(2:5, names_to = "PC") 

df_pc2 <- data.frame(t = mid, fpca_mod$efunctions) %>%
  rename(PC1=X1, PC2=X2, PC3=X3, PC4=X4) %>%
  pivot_longer(2:5, names_to = "PC") 

left_join(df_pc1, df_pc2, by = c("t", "PC")) %>%
  ggplot()+
  geom_line(aes(x=t, y=value.x, col = PC), linewidth = 0.2)+
  geom_point(aes(x=t, y=value.y, col = PC), na.rm = T, alpha = 0.5, size = 0.5)+
  labs(x="Time", y="", title = "Re-evaluate eigenfunctions on the original grid")

```


## Debias the eigenvalues with GLMM


```{r revaluation}
df_phi <- data.frame(sind = 1:J, df_phi)
colnames(df_phi) <- c("sind", paste0("phi", 1:K))
train_df <- train_df %>% 
  left_join(df_phi, by = "sind")
train_df$id <- as.factor(train_df$id)

t1 <- Sys.time()
debias_glmm <- bam(Y ~ s(sind, bs="cc")+
                     s(id, by=phi1, bs="re")+
                     s(id, by=phi2, bs="re")+
                     s(id, by=phi3, bs="re")+
                     s(id, by=phi4, bs="re"), 
                   family = binomial, 
                   data=train_df, 
                   method = "fREML",
                   discrete = TRUE)
t2 <- Sys.time()
# t2-t1
```

I ran into a few problems at this step. Not only does the mgcv::bam model take a really long time to fit, the re-evaluated eigenvalues are not proportional to the FPCA eigenvalues. They also lost the decreasing nature. Specifically:

- The smoothing parameters from the mgcv::bam model: `r debias_glmm$sp[2:5]`
- The re-evaluated eigenvalues, which are inverse of the smoothing parameters above: `r 1/debias_glmm$sp[2:5]`
- The **biased** eigenvalues from FPCA (step 3): `r fpca_mod$evalues[1:4]`

The FPCA eigenvalues, as expected, decreases from PC1 to PC4. However after re-evaluation, the eigenvalue corresponding to PC2 is less than PC3. Also, the proportion of these two set of estimates is not constant, as below: 

```{r, class.source = 'fold-show'}
# proportion of biased eigenvalues to re-evaluated eigenvalues
fpca_mod$evalues/(1/debias_glmm$sp[2:5])
```







