---
title: "Progress Report"
author: "Ying Jin"
date: "2023-05-16"
output: 
  html_document:
    self_contained: no
    number_sections: true
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

set.seed(516)

library(here)
library(tidyverse)
library(ggpubr)
library(refund)
library(lme4)
library(ROCR)
library(GLMMadaptive)
theme_set(theme_minimal())

df <- read_rds(here("Data/nhanes_bi.rds"))
load(here("Data/ApplOutput_fGFPCA.RData"))

```


# NHANES data application

## Data overview

- 8763 subjects, 1440 measures each
- no missingness

```{r}
N <- length(unique(df$SEQN)) # sample size 8763
J <- max(df$sind) # 1440 measures per subject
```


```{r}
rand_id <- sample(unique(df$SEQN), size = 4) # "toy" sample

df %>% 
  filter(SEQN %in% rand_id) %>%
  ggplot()+
  geom_point(aes(x=sind, y=Z), size = 0.5)+
  facet_wrap(~SEQN)+
  labs(x="Time", y = "Activity", title = "A brief overview of the outcome")
```

## fGFPCA Estimation

```{r}
# code
source(here("Code/GLMM-FPCA.R")) 
# use pred_latent function to estimate latent function 
source(here("Code/OutSampMLE.R"))
# source(here("Code/OutsampBayes.R"))


```

### Binning

- Bin every 10 measures 

```{r, message=FALSE}
# bin data
bin_w <- 10 # bin width
n_bin <- J/bin_w # number of bins
brks <- seq(0, J, by = bin_w) # cutoff points
mid <- (brks+bin_w/2)[1:n_bin] # mid points


df$bin <- cut(df$sind, breaks = brks, include.lowest = T, labels = mid)
df$bin <- as.numeric(as.character(df$bin))
# head(df)

df %>% 
  filter(SEQN %in% rand_id) %>%
  group_by(SEQN, bin) %>%
  summarise(num = sum(Z)) %>%
  ggplot()+
  geom_point(aes(x=bin, y=num), size = 0.5)+
  facet_wrap(~SEQN)+
  labs(x="Time", y = "Activity", title = "Number of active nimutes within each bin")
```


### Local GLMMs

- This is where near-unidentifiability issues first appear
- Though point estimation not necessarily affected


```{r, cache=TRUE, warning=FALSE}
# split by each bin
df <- df %>% rename(id = SEQN, Y=Z)
df_bin_lst <- split(df, f = df$bin)

# fit local GLMM and estimate latent function
# near-unidentifiability issues 
df_est_latent <- lapply(df_bin_lst, function(x){pred_latent(x)}) 
df_est_latent <- bind_rows(df_est_latent)

# keep midpoint values
df_est_latent <- df_est_latent %>%
  select(-sind, -Y) %>% distinct(.)
```


```{r, message=FALSE}
df %>% 
  filter(id %in% rand_id) %>%
  left_join(df_est_latent, by = c("id", "bin")) %>%
  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  ggplot()+
  geom_line(aes(x=sind, y=eta_hat, group = id))+
  geom_point(aes(x=sind, y = Y, group = id), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  geom_vline(xintercept = c(465, 545, 555, 1185, 1195, 1205), col = "red",
             alpha = 0.5, linewidth = 0.5)+
  labs(x = "Time", y = "Estimated latent function (probablity scale)")
```


```{r, message=FALSE}
# overview at estimated latent function
df %>% 
  filter(id %in% rand_id) %>%
  group_by(id, bin) %>%
  summarise(num = sum(Y)) %>%
  left_join(df_est_latent, by = c("id", "bin")) %>%
  # mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_hat, group = id))+
  geom_point(aes(x=bin, y = num, group = id), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  geom_vline(xintercept = c(465, 545, 555, 1185, 1195, 1205), col = "red",
             linewidth = 0.5, alpha = 0.6)+
  labs(x = "Bin", y = "Estimated latent function/number of active minutes")
```

#### Take a closer look at the unidentifiable bins

```{r, eval=FALSE}
try_fit_lst<-list()
for(i in seq_along(df_bin_lst)){
  this_glm <- tryCatch({glmer(Y ~ 1 + (1|id), data = df_bin_lst[[i]],
                             family = binomial)},
                       warning=function(w){NA})
  try_fit_lst[[i]] <- this_glm
  
}

# bins where glmer returned errors
names(df_bin_lst)[is.na(try_fit_lst)]
```

The fitting problem occurs in the following bins: 465, 545, 555, 1185, 1195, 1205. All of them had two warnings: 1) Non-convergence; 2) near-unidenetifiability. Summary results show that standar error estimates of fixed effects (intercept) are very close two zero. 

Below is an example of bin 465:

```{r class.source='fold-show'}
issue_fit <- glmer(Y ~ 1 + (1|id), data = df_bin_lst[['465']], family = binomial)

summary(issue_fit)
```


However, this issue seems to disappear after changing nAGQ. This argument controls the number of nodes in adaptive Gauss-Hermite quadrature approximation. If nAGQ=0, optimitization is done through penalized iteratively reweighted least square (PIRLS); if nAGQ=1, Laplace approximation and if nAGQ>1, adaptive Gauss-Hermite Quadrature. 

Reference: https://arxiv.org/pdf/2202.07864.pdf (In fact based on this reference the recommended number of node is 0)

Another reference: https://stats.stackexchange.com/questions/544937/when-is-it-appropriate-to-set-nagq-0-in-glmer. According to this one, when setting nAGQ=0, fixed effect parameters are not optimized.

For example set nAGQ=2 for bin 465: 

```{r class.source='fold-show'}
issue_fit <- glmer(Y ~ 1 + (1|id), data = df_bin_lst[['465']], family = binomial, nAGQ=2)

summary(issue_fit)
```


<!-- https://www.learn-mlms.com/07-module-7.html -->

<!-- https://rstudio-pubs-static.s3.amazonaws.com/33653_57fc7b8e5d484c909b615d8633c01d51.html -->

So I decided to re-do latent function estimation while adjusting the nAGQ parameter. Below are some findings:

- nAGQ=0 for all bins: no warnings appeared. This is supposed to be fast but not accurate. 
- nAGQ=1, 2, 3: warnings appeared in some bins. 
- nAGQ=5: no warnings! 

```{r, eval=FALSE}
# re-fit local GLMM and check numeric issues
try_fit_lst<-list()
for(i in seq_along(df_bin_lst)){
  this_glm <- tryCatch({glmer(Y ~ 1 + (1|id), data = df_bin_lst[[i]], family = binomial, nAGQ=5)},
                       warning=function(w){NA})
  try_fit_lst[[i]] <- this_glm
  
}

# bins where glmer returned errors
names(df_bin_lst)[is.na(try_fit_lst)]
glmer(Y ~ 1 + (1|id), data = df_bin_lst[['565']], family = binomial, nAGQ=3)
```

So perhaps we could try either setting nAGQ = 0 or a variable nAGQ:

Below compares the result for nAGQ = 0 and nAGQ = 5: 


```{r, cache=TRUE}
# re-define pre_latent funciton with nAGQ = 0
pred_latent_0AGQ <- function(df){
    this_glm <- glmer(Y ~ 1 + (1|id), data = df, family = binomial, nAGQ = 0)
    eta_hat <- predict(this_glm, type = "link")
    df$eta_hat <- eta_hat
    return(df)
}

# re-fit local GLMMs
t1=Sys.time()
df_est_latent_0AGQ <- lapply(df_bin_lst, function(x){pred_latent_0AGQ(x)}) 
t2=Sys.time()
# head(df_est_latent_0AGQ[[1]])
df_est_latent_0AGQ <- bind_rows(df_est_latent_0AGQ)
t_0AGQ <- t2-t1

# keep midpoint values
df_est_latent_0AGQ <- df_est_latent_0AGQ %>%
  select(-sind, -Y) %>% distinct(.)
```



```{r, cache=TRUE}
# re-define pre_latent funciton with nAGQ = 0
pred_latent_5AGQ <- function(df){
    this_glm <- glmer(Y ~ 1 + (1|id), data = df, family = binomial, nAGQ = 5)
    eta_hat <- predict(this_glm, type = "link")
    df$eta_hat <- eta_hat
    return(df)
}

# re-fit local GLMMs
t1=Sys.time()
df_est_latent_5AGQ <- lapply(df_bin_lst, function(x){pred_latent_5AGQ(x)}) 
t2=Sys.time()
# head(df_est_latent_0AGQ[[1]])
df_est_latent_5AGQ <- bind_rows(df_est_latent_5AGQ)
t_5AGQ <- t2-t1

# keep midpoint values
df_est_latent_5AGQ <- df_est_latent_5AGQ %>%
  select(-sind, -Y) %>% distinct(.)
```

- Time spent: `r t_0AGQ` for nAGQ=0, `r t_5AGQ` for nAGQ=5



```{r, message=FALSE}
ggarrange(
# nAQG=0
df %>% 
  filter(id %in% rand_id) %>%
  left_join(df_est_latent_0AGQ, by = c("id", "bin")) %>%
  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  ggplot()+
  geom_line(aes(x=sind, y=eta_hat, group = id))+
  geom_point(aes(x=sind, y = Y, group = id), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  # geom_vline(xintercept = c(465, 545, 555, 1185, 1195, 1205), col = "red",
             # alpha = 0.5, linewidth = 0.5)+
  labs(x = "Time", y = "Estimated latent function (probablity scale)", title = "nAGQ=0"),
#nAGQ=5
df %>% 
  filter(id %in% rand_id) %>%
  left_join(df_est_latent_5AGQ, by = c("id", "bin")) %>%
  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  ggplot()+
  geom_line(aes(x=sind, y=eta_hat, group = id))+
  geom_point(aes(x=sind, y = Y, group = id), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  # geom_vline(xintercept = c(465, 545, 555, 1185, 1195, 1205), col = "red",
             # alpha = 0.5, linewidth = 0.5)+
  labs(x = "Time", y = "Estimated latent function (probablity scale)", title = "nAGQ=5"), nrow = 1)
```



```{r, message=FALSE}
# overview at estimated latent function
ggarrange(
#nAGQ=0
df %>% 
  filter(id %in% rand_id) %>%
  group_by(id, bin) %>%
  summarise(num = sum(Y)) %>%
  left_join(df_est_latent_0AGQ, by = c("id", "bin")) %>%
  # mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_hat, group = id))+
  geom_point(aes(x=bin, y = num, group = id), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  # geom_vline(xintercept = c(465, 545, 555, 1185, 1195, 1205), col = "red",
  #            linewidth = 0.5, alpha = 0.6)+
  labs(x = "Bin", y = "Estimated latent function/number of active minutes", title = "nAGQ=0")+
  ylim(-25, 10),

#nAGQ=5
df %>% 
  filter(id %in% rand_id) %>%
  group_by(id, bin) %>%
  summarise(num = sum(Y)) %>%
  left_join(df_est_latent_5AGQ, by = c("id", "bin")) %>%
  # mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_hat, group = id))+
  geom_point(aes(x=bin, y = num, group = id), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  # geom_vline(xintercept = c(465, 545, 555, 1185, 1195, 1205), col = "red",
  #            linewidth = 0.5, alpha = 0.6)+
  labs(x = "Bin", y = "Estimated latent function/number of active minutes", title = "nAGQ=5")+
  ylim(-25, 10),

nrow = 1
)
```

Obviously the lower bound of estimated latent function got a lot lower using AGQ than PIRLS. Nothing else seems to change much. 


### FPCA

I would also like to compare the difference between AGQ (nAGQ=5) and PIRLS (nAGQ=0)

- AGQ has 16 eigenfunctions while PIRLS has 18


```{r}
# nAGQ=0
mat_est_unique_0AGQ <- matrix(df_est_latent_0AGQ$eta_hat, nrow=N, ncol=n_bin, byrow = F) 
# row index subject, column binned time
fpca_mod_0AGQ <- fpca.face(mat_est_unique_0AGQ, pve = 0.95, argvals = mid, var=T)

K_0AGQ <- ncol(fpca_mod_0AGQ$efunctions) # 18 eigenfunctions
```


```{r}
# nAGQ=5
mat_est_unique_5AGQ <- matrix(df_est_latent_5AGQ$eta_hat, nrow=N, ncol=n_bin, byrow = F) 
# row index subject, column binned time
fpca_mod_5AGQ <- fpca.face(mat_est_unique_5AGQ, pve = 0.95, argvals = mid, var=T)

K_5AGQ <- ncol(fpca_mod_5AGQ$efunctions) # 18 eigenfunctions
```

- Population mean function

```{r}
# population mean
par(mfrow=c(1, 2))
plot(mid, fpca_mod_0AGQ$mu, type = "l", xlab = "bin", ylab = "mean", main = "nAGQ=0", ylim = c(-25, 0))
plot(mid, fpca_mod_5AGQ$mu, type = "l", xlab = "bin", ylab = "mean", main = "nAGQ=5", ylim = c(-25, 0))
```
- PC functions
```{r, fig.height=10, fig.width=6}
# eigenfunctions
par(mfrow=c(4,2))

plot(mid, fpca_mod_0AGQ$efunctions[, 1], type="l", xlab="bin", ylab="PC1", main = "nAGQ=0")
plot(mid, fpca_mod_5AGQ$efunctions[, 1], type="l", xlab="bin", ylab="PC1", main = "nAGQ=5")

plot(mid, fpca_mod_0AGQ$efunctions[, 2], type="l", xlab="bin", ylab="PC2")
plot(mid, fpca_mod_5AGQ$efunctions[, 2], type="l", xlab="bin", ylab="PC3")

plot(mid, fpca_mod_0AGQ$efunctions[, 3], type="l", xlab="bin", ylab="PC3")
plot(mid, fpca_mod_5AGQ$efunctions[, 3], type="l", xlab="bin", ylab="PC3")

plot(mid, fpca_mod_0AGQ$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
plot(mid, fpca_mod_5AGQ$efunctions[, 4], type="l", xlab="bin", ylab="PC4")
```

- Individual covariance matrix
```{r}
# covariance
heatmap(fpca_mod_0AGQ$VarMats[[which(unique(df_est_latent$id)==rand_id[1])]],
        Rowv = NA, Colv = NA, main = "nAGQ=0")
heatmap(fpca_mod_5AGQ$VarMats[[which(unique(df_est_latent$id)==rand_id[1])]],
        Rowv = NA, Colv = NA, main = "nAGQ=5")
```

Compared the nAGQ=1: 

- The lower bound of mean function went up
- Shapes of PC functions didn't change a lot
- Covariance matrix looks a lot more different! Near-diagonal correlation got higher and off-diagonal got lower.

## Dynamic prediction

### Maximum likelihood

- Boundary hitting issues with numeric maximization
- Warning message: Steady solution not reached
- Large variation in both predicted values and scores. The numeric solution from maximum likelihood is obviously not stable. It is very likely to fail to converge.
- Perhaps it can be fixed with numeric options? But I'd rather not go that route because it depend on chance a lot. 
- After setting nAGQ=0, the problem got better for some subjects, but not the others. 


```{r}
# Out-of-sample prediction
# maximum observations time: 360, 720, 1080
# nAGQ=0
df_pred <- df_est_latent_0AGQ %>% filter(id %in% rand_id)
df_pred[, 'pred_t360'] <- df_pred[, 'pred_t720'] <- df_pred[, 'pred_t1080'] <- NA
score_out_mat <- array(NA, dim = c(length(rand_id), K_0AGQ, 3))
# dim indexes subject, eigenfunction and max obs time respectively

# nAGQ=5
df_pred2 <- df_est_latent_5AGQ %>% filter(id %in% rand_id)
df_pred2[, 'pred_t360'] <- df_pred2[, 'pred_t720'] <- df_pred2[, 'pred_t1080'] <- NA
score_out_mat2 <- array(NA, dim = c(length(rand_id), K_5AGQ, 3))
```

```{r, warning=FALSE}
# prediction for a single subject
# nAGQ=0
for(i in rand_id){
  df_i <- df %>% filter(id==i) %>% mutate(bin=as.numeric(bin))
  # prediction
  pred_t360 <- out_samp_dyn_pred(df_new = df_i %>% filter(sind<=360),
                                 fpca_fit = fpca_mod_0AGQ, K = K_0AGQ)
  pred_t720 <- out_samp_dyn_pred(df_new = df_i %>% filter(sind<=720),
                                 fpca_fit = fpca_mod_0AGQ, K = K_0AGQ)
  pred_t1080 <- out_samp_dyn_pred(df_new = df_i %>% filter(sind<=1080),
                                  fpca_fit = fpca_mod_0AGQ, K = K_0AGQ)
  # prediction in container
  df_pred[df_pred$id==i, 'pred_t360'] <- pred_t360$eta_pred
  df_pred[df_pred$id==i, 'pred_t720'] <- pred_t720$eta_pred
  df_pred[df_pred$id==i, 'pred_t1080'] <- pred_t1080$eta_pred
  # score in container
  # score_out_mat[i, ,1] <- pred_t360$score_out
  # score_out_mat[i, ,2] <- pred_t720$score_out
  # score_out_mat[i, ,3] <- pred_t1440$score_out
}

# df_pred

```


```{r, warning=FALSE}
# prediction for a single subject
# nAGQ=5
for(i in rand_id){
  df_i <- df %>% filter(id==i) %>% mutate(bin=as.numeric(bin))
  # prediction
  pred_t360 <- out_samp_dyn_pred(df_new = df_i %>% filter(sind<=360),
                                 fpca_fit = fpca_mod_5AGQ, K = K_5AGQ)
  pred_t720 <- out_samp_dyn_pred(df_new = df_i %>% filter(sind<=720),
                                 fpca_fit = fpca_mod_5AGQ, K = K_5AGQ)
  pred_t1080 <- out_samp_dyn_pred(df_new = df_i %>% filter(sind<=1080),
                                  fpca_fit = fpca_mod_5AGQ, K = K_5AGQ)
  # prediction in container
  df_pred2[df_pred2$id==i, 'pred_t360'] <- pred_t360$eta_pred
  df_pred2[df_pred2$id==i, 'pred_t720'] <- pred_t720$eta_pred
  df_pred2[df_pred2$id==i, 'pred_t1080'] <- pred_t1080$eta_pred
  # score in container
  # score_out_mat[i, ,1] <- pred_t360$score_out
  # score_out_mat[i, ,2] <- pred_t720$score_out
  # score_out_mat[i, ,3] <- pred_t1440$score_out
}

# df_pred

```

```{r}
df_pred$pred_t360[df_pred$bin<=360] <- NA
df_pred$pred_t720[df_pred$bin<=720] <- NA
df_pred$pred_t1080[df_pred$bin<=1080] <- NA
# df_pred %>% filter(id == rand_id[1])

df_pred2$pred_t360[df_pred2$bin<=360] <- NA
df_pred2$pred_t720[df_pred2$bin<=720] <- NA
df_pred2$pred_t1080[df_pred2$bin<=1080] <- NA
```


```{r, warning=FALSE}
ggarrange(
df_pred %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_hat, group = id))+
  geom_line(aes(x=bin, y = pred_t360, col = "360", group = id))+
  geom_line(aes(x=bin, y = pred_t720, col = "720", group = id))+
  geom_line(aes(x=bin, y = pred_t1080, col = "1080", group = id))+
  facet_wrap(~id,scales = "free")+
  labs(title = "nAGQ=0"),

df_pred2 %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_hat, group = id))+
  geom_line(aes(x=bin, y = pred_t360, col = "360", group = id))+
  geom_line(aes(x=bin, y = pred_t720, col = "720", group = id))+
  geom_line(aes(x=bin, y = pred_t1080, col = "1080", group = id))+
  facet_wrap(~id,scales = "free")+
  labs(title = "nAGQ=5"),

nrow = 1
)
```


### Bayes method like Laplace approximation


```{r, warning=FALSE, message=FALSE}
#PIRLS
df_pred <- df_est_latent_0AGQ %>% filter(id %in% rand_id)
df_pred[, 'pred_t360'] <- df_pred[, 'pred_t720'] <- df_pred[, 'pred_t1080'] <- NA
score_out_mat <- array(NA, dim = c(length(rand_id), K_0AGQ, 3))

# AGQ
df_pred2 <- df_est_latent_5AGQ %>% filter(id %in% rand_id)
df_pred2[, 'pred_t360'] <- df_pred2[, 'pred_t720'] <- df_pred2[, 'pred_t1080'] <- NA
score_out_mat <- array(NA, dim = c(length(rand_id), K_5AGQ, 3))

```


```{r, results='hide'}
# unique id
# prediction for a single subject
source(here("Code/OutsampBayes.R"))

# PRILS
for(i in rand_id){
  df_i <- df %>% filter(id==i) 
  
  # prediction 
  ## up to 360
  pred_t1 <- out_pred_laplace(fpca_mod_0AGQ, df_i %>% filter(sind<=360))
  eta_pred_t1 <- fpca_mod_0AGQ$mu+fpca_mod_0AGQ$efunctions%*%pred_t1$score_out
  df_pred[df_pred$id == i, 'pred_t360'] <- eta_pred_t1
  ## up to 720
  pred_t2 <- out_pred_laplace(fpca_mod_0AGQ, df_i %>% filter(sind<=720))
  eta_pred_t2 <- fpca_mod_0AGQ$mu+fpca_mod_0AGQ$efunctions%*%pred_t2$score_out
  df_pred[df_pred$id == i, 'pred_t720'] <- eta_pred_t2
  ## up to 1080
  pred_t3 <- out_pred_laplace(fpca_mod_0AGQ, df_i %>% filter(sind<=1080))
  eta_pred_t3 <- fpca_mod_0AGQ$mu+fpca_mod_0AGQ$efunctions%*%pred_t3$score_out
  df_pred[df_pred$id == i, 'pred_t1080'] <- eta_pred_t3

}

df_pred$pred_t360[df_pred$bin<=360] <- NA
df_pred$pred_t720[df_pred$bin<=720] <- NA
df_pred$pred_t1080[df_pred$bin<=1080] <- NA
```


```{r, results='hide'}
# AGQ
for(i in rand_id){
  df_i <- df %>% filter(id==i) 
  
  # prediction 
  ## up to 360
  pred_t1 <- out_pred_laplace(fpca_mod_5AGQ, df_i %>% filter(sind<=360))
  eta_pred_t1 <- fpca_mod_5AGQ$mu+fpca_mod_5AGQ$efunctions%*%pred_t1$score_out
  df_pred2[df_pred2$id == i, 'pred_t360'] <- eta_pred_t1
  ## up to 720
  pred_t2 <- out_pred_laplace(fpca_mod_5AGQ, df_i %>% filter(sind<=720))
  eta_pred_t2 <- fpca_mod_5AGQ$mu+fpca_mod_5AGQ$efunctions%*%pred_t2$score_out
  df_pred2[df_pred2$id == i, 'pred_t720'] <- eta_pred_t2
  ## up to 1080
  pred_t3 <- out_pred_laplace(fpca_mod_5AGQ, df_i %>% filter(sind<=1080))
  eta_pred_t3 <- fpca_mod_5AGQ$mu+fpca_mod_5AGQ$efunctions%*%pred_t3$score_out
  df_pred2[df_pred2$id == i, 'pred_t1080'] <- eta_pred_t3

}

df_pred2$pred_t360[df_pred2$bin<=360] <- NA
df_pred2$pred_t720[df_pred2$bin<=720] <- NA
df_pred2$pred_t1080[df_pred2$bin<=1080] <- NA
# df_pred %>% filter(id == rand_id[1])
```

```{r, warning=FALSE}
ggarrange(
df_pred %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_hat, group = id))+
  geom_line(aes(x=bin, y = pred_t360, col = "360", group = id))+
  geom_line(aes(x=bin, y = pred_t720, col = "720", group = id))+
  geom_line(aes(x=bin, y = pred_t1080, col = "1080", group = id))+
  facet_wrap(~id,scales = "free")+
  labs(title = "nAGQ=0"),
  
df_pred2 %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_hat, group = id))+
  geom_line(aes(x=bin, y = pred_t360, col = "360", group = id))+
  geom_line(aes(x=bin, y = pred_t720, col = "720", group = id))+
  geom_line(aes(x=bin, y = pred_t1080, col = "1080", group = id))+
  facet_wrap(~id,scales = "free")+
  labs(title = "nAGQ=5"), nrow = 1
)
```






<!-- ## Failed prediction  -->

<!-- - Laplace approximation failed for some subjects -->
<!-- - Only when the observed track is short (up to 360) -->
<!-- - These subjects have something in common -->

<!-- ```{r} -->
<!-- # compare subjects that failed at Laplace Approximation with others -->
<!-- rand_id <- sample(unique(df$SEQN), size = 4) -->
<!-- com_id <- c(rand_id, sample(skip_id, 4)) -->

<!-- df_est_latent %>% filter(id %in% com_id) %>% -->
<!--   left_join(df %>% select(SEQN, Z, sind) %>% -->
<!--               rename(id = SEQN, Y=Z, bin = sind), by = c("id", "bin")) %>% -->
<!--   mutate(type = ifelse(id %in% rand_id, "Prediction", "Failed prediction")) %>% -->
<!--   ggplot(aes(x=bin, y=eta_hat, col = type))+ -->
<!--   geom_line()+ -->
<!--   geom_vline(xintercept = 360)+ -->
<!--   facet_wrap(~type+id, nrow = 2, ncol = 4)+ -->
<!--   labs(y = "Estimated latend function", x = "Time", col = "") -->
<!-- ``` -->


<!-- ## Subjects with complete prediction -->

<!-- ```{r fig_prob, warning=FALSE} -->
<!-- df_est_latent %>% -->
<!--   filter(id %in% rand_id) %>% -->
<!--   mutate(pred_t1080 = ifelse(bin<=1080, NA, pred_t1080), -->
<!--          pred_t720 = ifelse(bin<=720, NA, pred_t720), -->
<!--          pred_t360 = ifelse(bin<=360, NA, pred_t360)) %>% -->
<!--   mutate_at(vars(pred_t360, pred_t720, pred_t1080), function(x)exp(x)/(1+exp(x))) %>% -->
<!--   left_join(df %>% select(SEQN, Z, sind) %>% -->
<!--               rename(id = SEQN, Y=Z, bin = sind), by = c("id", "bin")) %>% -->
<!--   ggplot()+ -->
<!--     geom_line(aes(x=bin, y = pred_t360, col = "360"), linetype = "dashed")+ -->
<!--     geom_line(aes(x=bin, y = pred_t720, col = "720"),linetype = "dashed")+ -->
<!--     geom_line(aes(x=bin, y = pred_t1080, col = "1080"),linetype = "dashed")+ -->
<!--     geom_point(aes(x=bin, y = Y, col = "Outcome"), size = 0.5)+ -->
<!--     facet_wrap(~id)+ -->
<!--     labs(x = "Time", y = "Estimated latent function (probablity scale)", -->
<!--          col = "Maximum observation time") -->
<!-- ``` -->


<!-- ```{r fig_eta, warning=FALSE} -->
<!-- df_est_latent %>% -->
<!--   filter(id %in% rand_id) %>% -->
<!--   mutate(pred_t1080 = ifelse(bin<=1080, NA, pred_t1080), -->
<!--          pred_t720 = ifelse(bin<=720, NA, pred_t720), -->
<!--          pred_t360 = ifelse(bin<=360, NA, pred_t360)) %>% -->
<!--   ggplot()+ -->
<!--     geom_line(aes(x=bin, y = pred_t360, col = "360"), linetype = "dashed")+ -->
<!--     geom_line(aes(x=bin, y = pred_t720, col = "720"),linetype = "dashed")+ -->
<!--     geom_line(aes(x=bin, y = pred_t1080, col = "1080"),linetype = "dashed")+ -->
<!--     geom_line(aes(x=bin, y = eta_hat, col = "eta_hat"), linetype = "solid" )+ -->
<!--     facet_wrap(~id)+ -->
<!--     labs(x = "Time", y = "Estimated latent function") -->
<!-- ``` -->

<!-- ## AUC on subjects with complete prediction -->

<!-- - Use interpolation to fill in gaps between bins -->

<!-- ```{r} -->
<!-- # subjects with complete prediction -->
<!-- N <- length(unique(df$SEQN))-length(skip_id) -->
<!-- eval_id <- setdiff(unique(df$SEQN), skip_id) -->
<!-- J <- max(df$sind) -->

<!-- # containter: row index subject, column index time -->
<!-- eta_mat_t360 <- eta_mat_t720 <- eta_mat_t1080 <- matrix(NA, nrow = N, ncol = J) -->

<!-- # interpolation -->
<!-- for(i in seq_along(eval_id)){ -->
<!--    eta_mat_t360[i, ] <- approx(x=df_est_latent$bin[df_est_latent$id==eval_id[i]], -->
<!--                                y=df_est_latent$pred_t360[df_est_latent$id==eval_id[i]],  -->
<!--                                xout = 1:J)$y -->
<!--    eta_mat_t720[i, ] <- approx(x=df_est_latent$bin[df_est_latent$id==eval_id[i]], -->
<!--                                y=df_est_latent$pred_t720[df_est_latent$id==eval_id[i]],  -->
<!--                                xout = 1:J)$y -->
<!--    eta_mat_t1080[i, ] <- approx(x=df_est_latent$bin[df_est_latent$id==eval_id[i]], -->
<!--                                y=df_est_latent$pred_t1080[df_est_latent$id==eval_id[i]],  -->
<!--                                xout = 1:J)$y -->
<!--   } -->


<!-- ``` -->



<!-- ```{r} -->
<!-- # true outcome -->
<!-- Y_mat <- df$Z[df$SEQN %in% eval_id] -->
<!-- Y_mat <- matrix(Y_mat, nrow = N, ncol = J, byrow = T)  -->
<!-- ``` -->

<!-- ```{r AUC_fGFPCA} -->
<!-- # maximum observations time -->
<!-- t_vec <- c(360, 720, 1080) -->

<!-- # AUC container -->
<!-- auc_mat <- matrix(NA, 3, 3) -->
<!-- colnames(auc_mat) <- c("360", "720", "1080") -->
<!-- rownames(auc_mat) <- c("(360, 720]", "(720, 1080]", "(1080, 1440]") -->

<!-- # Calculate AUC -->

<!-- auc_mat[1, 1] <- performance(prediction(as.vector(eta_mat_t360[, 361:720]),  -->
<!--                                    as.vector(Y_mat[, 361:720])), -->
<!--                         "auc")@y.values[[1]] -->
<!-- auc_mat[2, 1] <- performance(prediction(as.vector(eta_mat_t360[, 721:1080]),  -->
<!--                                    as.vector(Y_mat[, 721:1080])), -->
<!--                         "auc")@y.values[[1]] -->
<!-- auc_mat[3, 1] <- performance(prediction(as.vector(eta_mat_t360[, 1081:1435]),  -->
<!--                                    as.vector(Y_mat[, 1081:1435])), -->
<!--                         "auc")@y.values[[1]] -->

<!-- auc_mat[2, 2] <- performance(prediction(as.vector(eta_mat_t720[, 721:1080]),  -->
<!--                                    as.vector(Y_mat[, 721:1080])), -->
<!--                         "auc")@y.values[[1]] -->
<!-- auc_mat[3, 2] <- performance(prediction(as.vector(eta_mat_t720[, 1081:1435]),  -->
<!--                                    as.vector(Y_mat[, 1081:1435])), -->
<!--                         "auc")@y.values[[1]] -->


<!-- auc_mat[3, 3] <- performance(prediction(as.vector(eta_mat_t1080[, 1081:1435]),  -->
<!--                                    as.vector(Y_mat[, 1081:1435])), -->
<!--                         "auc")@y.values[[1]] -->
<!-- auc_mat -->
<!-- ``` -->


<!-- ## Reference method -->

<!-- - GLMMadaptive? -->
<!-- - fcr on lowess smoothed track? -->
